### 2014-02-06 thesis-proposal/problem.tex
More progress on proposal

% !TEX root = omar-thesis-proposal.tex
\vspace{-25pt}
\section{Motivation}\label{motivation}
Specifying and implementing a programming language together with its supporting tools (collectively, a \emph{programming system}) that has sound theoretical foundations, helps users identify and fix errors as early as possible, supports a natural programming style, and performs well across diverse problem domains and hardware platforms remains a grand challenge in computing. Due to the increasing diversity and complexity of modern problem domains and hardware platforms, it has become clear that no small  collection of general-purpose constructs backed by a simple, all-purpose implementation can fully satisfy these criteria in all cases. Instead,  researchers and domain experts (collectively, \emph{providers}) continue to develop  specialized notations,  static semantics, dynamic semantics, implementation strategies, optimizations, run-time systems and tools (collectively, \emph{features})  designed to address these modern challenges in different circumstances. %, researchers and domain experts must continue to design and develop, in a principled manner, specialized notations, type systems, optimizations, run-time systems and tools (collectively, \emph{features}) to support developers as they work in increasingly complex problem domains and on modern hardware systems.

% Rather, researchers and domain experts must continue, in a principled manner, to specify and implement novel abstractions, provide alternative implementations of existing abstractions, supplement languages with useful tools and improve tools with new behaviors (collectively, realize new \emph{features}) that better satisfy these criteria by balancing the concerns relevant to particular application domains.

Ideally, a provider would develop and distribute such new features orthogonally, as libraries, so that client developers could granularly choose those that best satisfy their needs. Unfortunately this is often infeasible because libraries are vehicles for specifying a program in terms of a monolithic collection of existing features. From the perspective of a library, the language's syntax, type system and semantics are fixed in advance, the compiler and run-time system are ``black box'' implementations of this fixed language, and the other tools, like code editors and debuggers, operate according to domain-agnostic protocols also based only on a fixed language specification. As a result, providers of new system features must often take \emph{language-external approaches}, sometimes designing a new system altogether. We will argue that such approaches are problematic, and that taking them has led to an unnecessary gap between research and practice. In place of these approaches, we will advocate for \emph{language-integrated extensibility mechanisms}, and show how, by organizing new features around {types} enforcing certain constraints, these mechanisms can be made both safe and highly expressive. We call a type that introduces new features an \emph{active type} and programming systems organized around active types \emph{actively-typed programming systems}.
%But taking a \emph{language-internal approach} to implementing a feature is the most practical. If a feature can be realized by creatively using existing language constructs and distributed as a library, clients face fewer barriers to adoption because it is easy to integrate library-based features into existing projects gradually and granularly and they leverage well-understood and well-developed mechanisms.
% But taking this approach is often \emph{not} possible today %We call designs \emph{monolithic programming systems}.

%To realize a new abstraction or system behavior, such experts can consider either a \emph{language-internal approach}, where they work within an existing language and distribute their solutions as libraries, or a \emph{language-external approach}, where they create a new, distinct programming system (often centered around what has come to be called a new \emph{domain-specific language} \cite{dsl}) or extend an existing system by some mechanism that is not part of the language itself, such as an extension mechanism supported by a {particular} compiler, editor or other tool.
\subsection{Motivating Example: Regular Expressions}\label{regex}
To make the issue concrete, let us begin with a simple example. \emph{Regular expressions} are a widely-used mechanism for finding patterns in semi-structured strings (e.g. DNA sequences) \cite{Thompson:1968:PTR:363347.363387}. A programming system that included full compile-time and edit-time support for regular expressions might simultaneously provide features like these:

\begin{enumerate}
\item \textbf{Built-in syntax for pattern literals} (e.g. \cite{perlre}) so that malformed patterns result in intelligible \emph{compile-time} parsing errors. In languages lacking such literals, run-time errors relating to pattern syntax can occur, even in well-tested, deployed code \cite{spishak2012type}. 
\item \textbf{Typechecking logic} that ensures that key constraints related to regular expressions are not violated at compile-time:
	\begin{enumerate}
	\item only appropriate values are spliced into regular expressions, to avoid splicing errors and injection attacks \cite{owasp2013}
	\item out-of-bounds backreferences are not used \cite{spishak2012type}
	\item strings known to be in the language of a regular expression are given a type that tracks this information and ensures that string manipulation operations do not inadvertently lead to a string that is not in the assumed language \cite{fulton-thesis}.
	\end{enumerate}
When a type error is found, an intelligible error message is provided.
\item \textbf{Elaboration logic} that partially or fully compiles known regular expressions into the efficient internal representation that will be used by the regular expression matching engine (e.g., a finite automata \cite{Thompson:1968:PTR:363347.363387}) ahead of time. In most languages, this compilation step occurs at run-time, even if the pattern is fully known at compile-time, thereby introducing performance overhead into programs. If the developer is not careful to cache compiled representations, regular expressions used repeatedly in a program might be needlessly re-compiled on each use. %By performing this step ahead-of-time, these dangers can be avoided.
\item \textbf{Editor services} for quickly testing regular expression patterns against test strings,  referring to documentation or searching databases of common patterns (e.g. \cite{IntelliJRegexp}).
\end{enumerate}

%In a conventional \emph{monolithic} programming system, support for each of these features would need to be built into the language and tools. 
No system today has built-in support for all of the features enumerated above. Instead, libraries generally provide support for regular expressions by leveraging  general-purpose constructs. Unfortunately, it is impossible to fully encode the syntax and the specialized static and dynamic semantics described above in terms of general-purpose notations and abstractions like objects or inductive datatypes. Library providers have thus needed to compromise, typically by asking clients to provide regular expressions as strings, thereby deferring parsing, typechecking and elaboration to run-time, which introduces performance overhead and can lead to unanticipated run-time errors (as shown in \cite{spishak2012type}) and security vulnerabilities (due to injection attacks when user inputs are spliced into patterns, for example). Similarly, useful tools for working with regular expressions are rarely integrated into editors, and even more rarely in a way that facilitates their discovery and use directly when the developer is manipulating regular expressions. Tools that must be discovered independently and accessed externally are used less frequently  and can be more awkward than necessary \cite{Campbell:2008:DRT:1636642.1636651, Omar:2012:ACC:2337223.2337324}, leading to lower productivity \cite{Omar:2012:ACC:2337223.2337324}.

\subsection{Language-External Approaches}
\begin{figure}
\begin{center}
\includegraphics[scale=.48]{approaches.pdf}
\end{center}
\vspace{-20px}
\caption{\small (a) When taking a language-external approach, new features are packaged together into separate languages and tools, causing problems with compatibility (described in the text). (b) When taking a language-integrated approach, there is one extensible host language and the compile-time and edit-time logic governing new constructs is expressed within ``active'' libraries. It is critical that the extension mechanism  guarantees that extensions cannot interfere with one another, so that the compatibility problem is avoided.}
\label{approaches}
\end{figure}

When the semantics or implementation of a system must be extended to fully realize a new feature, as in the example above, providers typically take a \emph{language-external approach}, either by developing a new or derivative programming system (supported by \emph{language workbenches} \cite{erdweg2013state},  \emph{DSL frameworks} \cite{fowler2010domain} or \emph{compiler generators}), or by extending an existing system by some mechanism that is not part of the language itself, such as an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting meaning-preserving optimizations, should be considered a pernicious means for creating new languages. Many ``practical'' compilers, including \texttt{gcc}, \texttt{GHC} and \texttt{SML/NJ}, are guilty of this sin, meaning that some programs that seem to be written in C, Haskell or Standard ML are actually written in tool-specific derivatives of those languages. Language-internal mechanisms do not lead to such fragmentation.} or other tool. For example, a researcher interested in providing regexp-related features (let us refer to these collectively as \texttt{R}) might design a new language with built-in support for regular expressions, perhaps basing it on an existing language (containing some general-purpose constructs, \texttt{G}). A different researcher developing a new language-based parallel programming abstraction or implementation strategy (collectively, \texttt{P}) might  take the same strategy. A third researcher, developing an alternative parallel programming abstraction (collectively, \texttt{Q}) might again do the same. This results in a collection of distinct systems as diagrammed in Figure \ref{approaches}a.%This latter method couples the semantics of the feature to the implementation details of a particular tool. Because the use of one implementation entails a different semantics for the feature than another, the extended tool acts, \emph{de facto}, as a distinct system for our purposes.

Unfortunately, when providers of new features take language-external approaches like this, it causes  problems for clients. Features cannot be adopted individually, but instead are only available coupled to a collection of other unrelated  features (e.g. the incidental features of the newly-designed language and its associated tools). This makes adoption more costly when these incidental features are  not desirable or insufficiently developed, or when the features bundled with a different language or tool are simultaneously desirable. Evidence supports the prominence of this problem: developers prefer language-integrated parallel programming abstractions to library-based implementations if all else is equal \cite{cave2010comparing}, but library-based implementations are more widely adopted because ``parallel programming languages'' privilege only a few chosen  abstractions and implementation strategies. This is problematic because different parallel abstractions or implementation strategies are more appropriate in different situations \cite{Tasharofi:2013rc} and moreover,  parallel programming support is rarely the only concern relevant to client developers outside of a classroom setting. Regular expression support, for example, may be simultaneously desirable for processing large amounts of textual data in parallel, but using these features together in the same compilation unit when they have been implemented by language-external means is difficult or impossible. One must either use the system containing features \texttt{G+R} or \texttt{G+P}. There is no system containing both. %Similarly, a language and tools designed primarily to support regular expressions might make an interesting research project, but it would not be a suitable tool for writing large applications with more varied needs.

@@

Different application concerns can sometimes be separated into different components, each written using a suitable language and tooling. This has been called the \emph{language-oriented approach} to software development \cite{journals/stp/Ward94}. Unfortunately, this is insufficient: the interface between components written in different languages remains an issue. The specialized constructs particular to one language (e.g. futures in a parallel programming language) cannot always be safely and naturally expressed in terms of those available in another (e.g. a general-purpose language), so building a program out of components written in a variety of different languages is difficult whenever these are exposed at interface boundaries. Tool support is also lost when calling into a different language. We call this fundamental issue the \emph{client compatibility problem}: clients of a certain collection of features cannot always interface with clients of a different collection of features in a safe, performant and natural manner.

One strategy often taken by proponents of the language-oriented approach to partially address the compatibility problem is to  target an established intermediate language, such as the Java Virtual Machine (JVM) bytecode, and support a superset of its constructs. Scala \cite{200464/IC} and F\# \cite{pickering2007foundations} are examples of prominent general-purpose languages that have taken this approach, and most DSL frameworks also rely on this strategy. As indicated in Figure \ref{approaches}a, this only enables true client compatibility in one direction. While calling into the common language becomes straightforward, calls in the other direction, or between the languages sharing the common target, are still restricted by the semantics of the common language.

If new languages only include constructs that can already be expressed safely and reasonably naturally in the common language, then this approach can work well.
But many of the most innovative constructs found in modern languages (often, the features that justify their creation) are difficult to define in terms of existing constructs in ways that guarantee all necessary invariants are statically maintained and that do not require large amounts boilerplate code and run-time overhead. For example, the type system of F\# guarantees that null values cannot occur within F\# data structures, but maintaining this important invariant still requires run-time checks because the typing rules of F\# do not apply when F\# code is called from other languages on the Common Language Infrastructure (CLI) like C\#. The F\# type system also includes support for checking that units of measure are used correctly \cite{syme2012expert, kennedy1994dimension}, but this specialized invariant is left completely unchecked at language boundaries. 
%In Scala, traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}. 
In some cases, desirable features must be omitted entirely due to concerns about  interoperability. F\#, for example, is based on Ocaml, but due to the need for bidirectional interoperability with CLI languages, it cannot support features like polymorphic variants, modules or functors \cite{ocaml-manual}.
%\end{itemize}

\subsection{Language-Integrated Approaches}
We argue that, due to these problems, taking a language-external approach to realizing a new feature should be considered harmful and avoided whenever possible. The goal of the research being proposed here is to design \emph{language-integrated extension mechanisms} that give developers the ability to define in libraries new edit-time and compile-time features of the system that have previously been defined centrally\footnote{One might compare today's monolithic programming systems to  {centrally-planned} economies, whereas extensible\- systems more closely resemble modern market economies.} so that such language-external approaches are less frequently necessary. More specifically, we will show how control over aspects of \textbf{syntax}, \textbf{typechecking}, \textbf{elaboration} and \textbf{editor services} can be delegated to user-defined logic distributed in {libraries}, as illustrated in Figure \ref{approaches}b. 
Such libraries are called \emph{active libraries} because, rather than being passive clients of features already available in the system, they contain logic invoked by the system during development or compilation to provide new features  \cite{activelibraries}. Features implemented within active libraries can be imported as needed by the clients of libraries that rely on them, unlike features implemented by language-external means, seemingly avoiding the client compatibility problem.

We must proceed with caution, however: critical issues having to do with {safety} must be overcome before language-integrated extension mechanisms can be introduced into a programming system, because if too much control over such core aspects of the system is given  to developers, the system may become unreliable. 
%For example, an extension could weaken important metatheoretic guarantees previously provided by the system. 
Type safety, for example, may not hold if the static and dynamic semantics of the language can be modified or extended arbitrarily from within libraries. Furthermore, even if extensions can be shown not to cause such problems in isolation, there may still be conflicts between extensions that could weaken their semantics, leading to subtle problems that only appear at link-time. For example, if two active libraries introduce the same syntactic form but back it with differing (valid) semantics, the issue would only manifest itself when both libraries were imported somewhere within the same compilation unit. These kinds of safety issues have plagued previous attempts to design language-integrated extensibility mechanisms.% To prevent them, our mechanisms will organize  extension logic around types to guarantee that extensions are both safe in isolation and also safely composable in any combination.

>>>>

% !TEX root = omar-thesis-proposal.tex
\vspace{-25pt}
\section{Motivation}\label{motivation}
Specifying and implementing a programming language together with its supporting tools (collectively, a \emph{programming system}) that is built upon sound theoretical foundations, helps users identify and fix errors as early as possible, supports a natural programming style, and performs well across diverse problem domains and hardware platforms remains a grand challenge in computing. In view of this goal, researchers and domain experts (collectively, \emph{providers}) continue to develop new special-purpose syntax, static and dynamic semantics, implementation strategies, optimizations, run-time systems and tools (collectively, \emph{features})  designed to address these challenges in increasingly diverse contexts. Ideally, a provider would be able to develop and distribute these kinds of new features orthogonally, as libraries, so that client developers could granularly choose those that best satisfy their needs. Unfortunately this is often infeasible because from the perspective of a library, the language's syntax and semantics are fully specified in advance, the compiler and run-time system are ``black box'' implementations of this fixed specification, and the other tools, like code editors and debuggers, operate according to fixed protocols. Providers of new system features must, as a result, take \emph{language-external approaches}, often by deriving a new programming system altogether. This approach has been encouraged, historically, by the availability of tools like compiler generators and language workbenches. We will argue that these approaches are technically problematic and that taking them has led to an unnecessary gap between research and practice. In their place, we will develop \emph{language-integrated extensibility mechanisms} that decentralize control over several core aspects of the programming system. By organizing new features around {types} and constraining them appropriately, we aim to show that these mechanisms can guarantee safety and non-interference of extensions while also being highly expressive.
%But taking a \emph{language-internal approach} to implementing a feature is the most practical. If a feature can be realized by creatively using existing language constructs and distributed as a library, clients face fewer barriers to adoption because it is easy to integrate library-based features into existing projects gradually and granularly and they leverage well-understood and well-developed mechanisms.
% But taking this approach is often \emph{not} possible today %We call designs \emph{monolithic programming systems}.

%To realize a new abstraction or system behavior, such experts can consider either a \emph{language-internal approach}, where they work within an existing language and distribute their solutions as libraries, or a \emph{language-external approach}, where they create a new, distinct programming system (often centered around what has come to be called a new \emph{domain-specific language} \cite{dsl}) or extend an existing system by some mechanism that is not part of the language itself, such as an extension mechanism supported by a {particular} compiler, editor or other tool.
\subsection{Motivating Example: Regular Expressions}\label{regex}
To make the problem we aim to address concrete, we begin with a simple example that we will return to throughout this work. \emph{Regular expressions} are a widely-used abstraction for finding patterns in semi-structured strings (e.g. DNA sequences) \cite{Thompson:1968:PTR:363347.363387}. If a programming system wished to fully support regular expressions, it might simultaneously provide features like these:

\begin{enumerate}
\item \textbf{Built-in syntax for pattern literals} (e.g. \cite{perlre}) so that the cognitive load of using regular expressions is low and malformed patterns result in intelligible {compile-time} errors.
\item A \textbf{static and dynamic semantics} that ensures, at compile-time whenever possible, that key invariants related to regular expressions are maintained:
	\begin{enumerate}
	\item only appropriate values are spliced into regular expressions, to avoid splicing errors and injection attacks \cite{owasp2013}
	\item out-of-bounds backreferences are not used \cite{spishak2012type}
	\item strings known to be in the language of a regular expression are given a type that tracks this information to ensure that string manipulation operations do not inadvertently lead to a string that is not in an assumed language \cite{fulton-thesis}.
	\end{enumerate}
When a type error is found, an intelligible error message is provided.
%\item An \textbf{implementation} that partially or fully compiles known regular expressions into the efficient internal representation that will be used by the regular expression matching engine (e.g., a finite automata \cite{Thompson:1968:PTR:363347.363387}) ahead of time. In most languages, this compilation step occurs at run-time, even if the pattern is fully known at compile-time, thereby introducing performance overhead into programs. If the developer is not careful to cache compiled representations, regular expressions used repeatedly in a program might be needlessly re-compiled on each use. %By performing this step ahead-of-time, these dangers can be avoided.
\item \textbf{Editor services} that allow clients to interactively test regular expression patterns against test strings, refer to documentation or search for common patterns (e.g. \cite{IntelliJRegexp}).
\end{enumerate}

%In a conventional \emph{monolithic} programming system, support for each of these features would need to be built into the language and tools. 
No system today builds in support for all of the features enumerated above. Instead, libraries generally provide support for regular expressions by leveraging  general-purpose abstraction mechanisms. Unfortunately, it is impossible to fully encode the syntax and the specialized static and dynamic semantics described above in terms of general-purpose notations and abstractions. Library providers thus need to compromise, typically by asking clients to provide regular expressions as strings and deferring parsing, typechecking and compilation to run-time. This introduces performance overhead and can lead to unanticipated run-time errors (as shown in \cite{spishak2012type}) and security vulnerabilities (due to injection attacks when user inputs are spliced into patterns, for example) \cite{string-stuff}. Useful tools for working with regular expressions are rarely integrated into editors, and even more rarely in a way that facilitates their discovery and use directly when the developer is manipulating regular expressions. Tools that must be discovered independently and accessed externally are used infrequently \cite{emersonStudy} and can be more awkward than necessary \cite{Campbell:2008:DRT:1636642.1636651, Omar:2012:ACC:2337223.2337324}, leading to lower productivity \cite{emersonStudy, Omar:2012:ACC:2337223.2337324}.\todo{complete references}

\subsection{Language-External Approaches}\label{external-approaches}
\begin{figure}
\begin{center}
\includegraphics[scale=.48]{approaches.pdf}
\end{center}
\vspace{-20px}
\caption{\small (a) When taking a language-external approach, new features are packaged together into separate languages and tools, causing problems with compatibility (described in the text). (b) When taking a language-integrated approach, there is one extensible host language and the compile-time and edit-time logic governing new constructs is expressed within ``active'' libraries. It is critical that the extension mechanism  guarantees that extensions cannot interfere with one another, so that the compatibility problem is avoided.}
\label{approaches}
%\vspace{-10px}
\end{figure}

When the semantics or implementation of a system must be extended to fully realize a new feature, as in the example above, providers typically take a \emph{language-external approach}, either by developing a new or derivative programming system (supported by \emph{language workbenches} \cite{erdweg2013state},  \emph{DSL frameworks} \cite{fowler2010domain} or \emph{compiler generators} \cite{abc}), or by extending an existing system by some mechanism that is not part of the language itself, such as an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting semantics-preserving optimizations, should be considered a pernicious means for creating new languages. Many ``practical'' compilers, including \texttt{gcc}, \texttt{GHC} and \texttt{SML/NJ}, are guilty of this, meaning that some programs that seem to be written in C, Haskell or Standard ML are actually written in tool-specific derivatives of these languages. Language-integrated mechanisms do not lead to such fragmentation.} or other tool. For example, a researcher interested in providing regular expression related features (let us refer to these collectively as \texttt{R}) might design a new system with built-in support for these, perhaps basing it on an existing system containing some general-purpose features (\texttt{G}). A different researcher developing a new language-based parallel programming abstraction or implementation strategy (\texttt{P}) might  take the same strategy. A third researcher, developing an alternative parallel programming abstraction (\texttt{Q}) might again do the same. This results in a collection of distinct systems as diagrammed in Figure \ref{approaches}a. Unfortunately, when providers of new features take language-external approaches like this, it causes  problems for clients related to \textbf{orthogonality} and \textbf{client compatibility}, described below. %This latter method couples the semantics of the feature to the implementation details of a particular tool. Because the use of one implementation entails a different semantics for the feature than another, the extended tool acts, \emph{de facto}, as a distinct system for our purposes.

\paragraph{Orthogonality} Features implemented by language-external means cannot be adopted individually, but instead are only available coupled to a fixed collection of other features. This makes adoption more costly when these incidental features are  not desirable or insufficiently developed, or when the features bundled with a different language or tool are simultaneously desirable. That is, one must either use the system containing features \texttt{G+R}, \texttt{G+P} or \texttt{G+Q}. There is no system containing \texttt{G}, \texttt{R}, \texttt{P} and \texttt{Q} in other combinations, and merging the systems containing each separately can be non-trivial because there is no requirement that providers of new features use a common mechanism. Even in cases where a common mechanism has been used, there are serious interference and safety issues, as we will discuss at length throughout this thesis (cf. Sec. \ref{background}, below).

Recent evidence emphasizes the importance of this problem for driving research into practice: developers prefer language-integrated parallel programming abstractions to library-based implementations if all else is equal \cite{cave2010comparing}, but library-based implementations are more widely adopted because ``parallel programming languages'' privilege only a few chosen  abstractions and associated implementation strategies. This is problematic because different parallel programming abstractions or implementation strategies are more appropriate in different situations \cite{Tasharofi:2013rc}. Moreover,  parallel programming support is rarely the only concern relevant to client developers outside of a classroom setting. Regular expression support, for example, may be simultaneously desirable for processing large amounts of textual data in parallel, but using these features together in the same compilation unit when they have been implemented by language-external means is difficult or impossible.  %Similarly, a language and tools designed primarily to support regular expressions might make an interesting research project, but it would not be a suitable tool for writing large applications with more varied needs.

@@

\paragraph{Client Compatibility} Even in cases where for each component of a system, there has been constructed a  system that is completely satisfactory, there remain problems at the interface between components. An interface that exposes the specialized constructs particular to one language (e.g. futures in a parallel programming language) cannot necessarily be safely and naturally consumed from another language (e.g. a general-purpose language). Tool support is also lost when calling into a different language. We call this fundamental issue the \emph{client compatibility problem}: code written by clients of a certain collection of features cannot always interface with code written by clients of a different collection  in a safe, performant and natural manner.

One strategy often taken by proponents of a \emph{language-oriented approach} to software development \cite{journals/stp/Ward94} to partially address the client compatibility problem is to  target an established intermediate language, such as the Java Virtual Machine (JVM) bytecode, and use its constructs as a common language for communication between components written in different languages. Scala \cite{200464/IC} and F\# \cite{pickering2007foundations} are examples of prominent general-purpose languages that have taken this approach, and most DSL frameworks also rely on this strategy. As indicated in Figure \ref{approaches}a, this only enables client compatibility in one direction. Calling into the common language becomes straightforward, but calling in the other direction, or between the languages sharing the common target, is not supported.

This approach can work well when new languages consist of constructs that can also be expressed safely and almost as naturally in the common language.
But many of the most innovative constructs found in modern languages (often, those that justify their creation) are difficult to define in terms of existing constructs in ways that guarantee all necessary invariants are statically maintained and that do not require large amounts boilerplate code and run-time overhead. For example, the type system of F\# guarantees that null values cannot occur within F\# data structures, but maintaining this important invariant still requires run-time checks because the typing rules of F\# do not apply when F\# code is called from other languages on the Common Language Infrastructure (CLI) like C\#. The F\# type system also includes support for checking that units of measure are used correctly \cite{syme2012expert, kennedy1994dimension}, but this highly-specialized invariant is left completely unchecked at language boundaries. 
%In Scala, traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}. 
In some cases, desirable features must be omitted entirely due to concerns about interoperability. F\#, for example, aimed to retain source compatibility with Ocaml code, but due to the need for bidirectional interoperability with CLI languages, it does not support features like polymorphic variants, modules or functors \cite{ocaml-manual} because they have no straightforward analogs in the type system of the CLI.
%\end{itemize}

\subsection{Language-Integrated Approaches}\label{language-integrated-approaches}
We argue that, due to these problems with orthogonality and client compatibility, taking a language-external approach to realizing a new feature should be considered harmful and avoided whenever possible. The goal of the research being proposed here is to design \emph{language-integrated extension mechanisms} that give providers the ability to define, within libraries, new features that have previously required central planning\footnote{One might compare today's programming systems to  {centrally-planned} economies, whereas extensible\- systems more closely resemble modern market economies. Our safety constraints serve a role analagous to market regulation. We leave further development of this analogy to the reader.} so that language-external approaches are less frequently necessary. More specifically, we will show how control over aspects of the \textbf{syntax}, \textbf{static and dynamic semantics} and \textbf{editor services} can be delegated to user-defined logic distributed in {libraries}, as illustrated in Figure \ref{approaches}b. 
Such libraries have been called \emph{active libraries}  \cite{activelibraries} because, rather than being passive clients of features already available in the system, they contain logic invoked by the system during development or compilation to provide new features. Features implemented within active libraries can be imported as needed, unlike features implemented by external means, seemingly avoiding the problems of orthogonality and client compatibility.

We must proceed with caution, however: critical issues having to do with {safety} must be overcome before language-integrated extension mechanisms can be introduced into a programming system. If too much control over  these core aspects of the system is given  to developers, the system may become unreliable. 
%For example, an extension could weaken important metatheoretic guarantees previously provided by the system. 
Type safety, for example, may not hold if the static and dynamic semantics of the language can be modified or extended arbitrarily from within libraries. Furthermore, even if extensions can be shown not to cause such problems in isolation, there may still be conflicts between extensions that could weaken their semantics, leading to subtle problems that only appear at link-time. For example, if two active libraries introduce the same syntactic form but back it with differing (but individually valid) semantics, the issue would only manifest itself when both libraries were imported somewhere within the same compilation unit. These kinds of safety issues have plagued previous attempts to design language-integrated extensibility mechanisms. We will briefly review some of these attempts below.% To prevent them, our mechanisms will organize  extension logic around types to guarantee that extensions are both safe in isolation and also safely composable in any combination.
